---
title: "Figure 5. Immune Cell Abundance by HPV"
author: "Caroline Wheeler"
date: "9/28/2021"
edits: "Malven - December 2023"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(reshape2)

knitr::opts_chunk$set(echo = TRUE)
```

read in CIBERSORT results
```{r}
cells <- read.csv("../data/CIBERSORTx_Job12_Results.csv") %>%
  select(-P.value, -Correlation, -RMSE)

clin <- read.csv("../data/file_id_BAM-expression_link.csv") %>%
  select(sample1, sample)
  
# Looks like this HPV prev table only contains 499 samples, but there are 500 in the files above 
hpv <- read.csv("../data/HPV-prevalence-per-sample.csv")
```

fix names
```{r}
cells <- cells %>% 
  mutate(sample1 = substr(Mixture, 2, length(Mixture))) %>%
  mutate(sample1 = gsub(".", "-", sample1, fixed = TRUE))
```

combine + clean
```{r}
cells <- merge(cells, clin)
cells <- merge(cells, hpv) %>%
  select(-sample1, -Mixture)
```

kruskal wallis test results in a table 
```{r}
results <- list()

for(i in colnames(cells[,2:23])){  
  results[[i]] <- kruskal.test(formula(paste(i, "~ HPV.prev")), data = cells)
}

results_table <- do.call(rbind.data.frame, results) %>%
  select(p.value) %>%
  dplyr::rename("p-value" = p.value)

collabs <- c("Naive B-cells", "Memory B-cells", "Plasma cells", "CD8+ T-cells", "Naive CD4 T-cells", "Memory CD4 T-cells, resting", "Memory CD4 T-cells, activated", "Follicular-helper T-cells", "Tregs", "Gamma delta T-cells", "NK cells, resting", "NK cells, activated", "Monocytes", "M0 Macrophages", "M1 Macrophages", "M2 Macrophages", "Dendritic cells, resting", "Dendritic cells, activated", "Mast cells, resting", "Mast cells, activated", "Eosinophils", "Neutrophils")

rownames(results_table) <- collabs
results
```

Write the full output to a table.
```{r}
results_table_join <- 
  results_table %>%
  rownames_to_column("name") %>%
  bind_cols(names(cells[2:23])) %>%
  dplyr::rename("cell" = "...3")

cells_median  <- cells %>%
  gather(-HPV.prev, -sample, key = "cell", value = "abundance") %>%
  group_by(HPV.prev, cell) %>%
  dplyr::summarize(median = median(abundance)) %>%
  spread(key = HPV.prev, value = median) %>%
  bind_cols(collabs) %>%
 full_join(results_table_join) %>%
  select("name", "0", "1", "p-value") %>%
  dplyr::rename("HPV-" = "0",
         "HPV+" = "1")

write.csv(cells_median,
          file = "../data/immune-cells_by_HPV_median_pvalue.csv", 
          row.names = FALSE)
```


subset those with a p-value < 0.05 and CD8+Tcells and Tregs and transform data
```{r}
cells <- cells %>%
  select(HPV.prev, B.cells.naive, T.cells.CD8, T.cells.follicular.helper, T.cells.regulatory..Tregs., Macrophages.M0, )

labs <- c("HPV.prev", "Naive B-cells", "CD8+ T-cells", "Follicular-helper T-cells", "Tregs", "M0 Macrophages")
colnames(cells) <- labs

cells <- melt(cells, id.vars=0:1)

cells <- cells %>%
  mutate(HPV.prev = ifelse(HPV.prev == 0, "HPV negative", "HPV postive"))
```

Boxplot to visualize
```{r}
# Figure 5 - Immune Cell Abundance of HPV+ and HPV-

bp <- ggplot(cells, aes(x=variable, y=value)) + geom_boxplot(aes(fill=HPV.prev)) + 
      xlab("") + ylab("Abundance") + labs(fill = "") + theme_bw(base_size = 12) + 
      theme(legend.key.size = unit(0.5, "cm"),
            legend.title = element_text(size = 11),
            legend.text = element_text(size = 10), 
            axis.text.x = element_text(angle = 45, hjust = 1)) 

bp

# save plot
ggsave(filename = "../figures/figure5_hnsc_immunecell_boxplot.png", plot = bp, width = 7, height = 6, dpi = 300)
```

