---
title: 'Survival: prevalence with covariates'
author: "Rebecca Hoyd"
date: "7/26/2021"
output: html_document

edited: "CW filter to subset of microbes and update paths"
date: "3/6/2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(survival)
#(survminer)
library(ggplot2)
library(tibble)
library(tidyr)

source("/fs/ess/PAS1695/exoticpipe/R/counts-to-rarefied-prevalence.R")
```

# Load data
```{r}
data.dir <- "/fs/ess/PAS1695/projects/HNSC/data/drake-output/7-15-2021"

exora <- read.csv(file.path(data.dir, "7-15-2021_tcga_exora-with-taxonomy.csv"), stringsAsFactors = F)
clin <- read.csv(file.path(data.dir, "7-15-2021_tcga_clinical.csv"), stringsAsFactors = F)

rcounts <- read.table("/fs/ess/PAS1695/generate-inputs_v1/TCGA-processing/data/HNSC/k2bout.txt",
                      stringsAsFactors = F, header = T, sep = "\t")

# taxa key
krakenmet = read.delim("/fs/ess/PAS1695/exoticpipe/external-data/kraken2-metaphlan-taxonomy.txt",
                         header = F, stringsAsFactors = F) %>%
    rename("Taxonomy" = "V1")
```

# Functions
```{r}
extract_coxph_results <- function(cp.form, int.term.name, dattab){
    tmp <- coxph(as.formula(cp.form), data = dattab) %>%
      summary()
  pvals <- tmp$coefficients %>%
    as.data.frame() %>%
    rownames_to_column(var = "term") %>%
    mutate(p.value = `Pr(>|z|)`) %>%
    select(term, p.value)
  conf.int <- tmp$conf.int %>%
    as.data.frame() %>%
    rownames_to_column(var = "term") %>%
    mutate(hazard.ratio = `exp(coef)`,
           confint.low = `lower .95`,
           confint.high = `upper .95`) %>%
    select(term, hazard.ratio, confint.low, confint.high) %>%
    left_join(pvals) %>%
    filter(term == int.term.name)
  return(conf.int)
}
```


I was going to  try to use a prevalence value from the normalized exora data, but the histogram looks a bit weird. Let's work with the raw counts instead.
```{r try working with exora}
prev.in <- exora %>%
  filter(microbe != "Homo.sapiens")
  
summary(prev.in$exo.ra)

prev.in %>%
  ggplot(aes(x = exo.ra)) +
  geom_histogram() +
  xlim(c(0, 1e-8))
```

```{r calculate rarefied prevalence SLOW}
# This looks like it's from the version of the pipeline that lost samples? we will ignore this I guess. I need to check with Caroline. Instead we'll use the rarefied prevalence on all available samples.
goodsamps <- intersect(exora$sample, clin$file_id.BAM)
length(goodsamps)

prevfun.in <- rcounts %>%
  column_to_rownames(var = "sample") %>%
  as.matrix()


# rarprev <- counts_to_rarefied_prevalenc(prevfun.in)
# 
# # Ask CW for write privilege
# write.csv(rarprev, "/fs/ess/PAS1695/projects/HNSC/data/rarefied-prevalence_all-TCGA-samps.csv",
#           row.names = F)
# write.csv(rarprev, "../data/rarefied-prevalence_all-TCGA-samps.csv",
#           row.names = F)

rarprev <- read.csv("/fs/ess/PAS1695/projects/HNSC/data/rarefied-prevalence_all-TCGA-samps.csv")
```

# Get all taxon levels, select genus level

# if this works just get rid of unclasside
```{r}
taxkey <- krakenmet %>%
  dplyr::filter(grepl("s__", Taxonomy)) %>%
  mutate(microbe = gsub(".*s__(.*)", "\\1", Taxonomy),
          microbe = make.names(microbe)) %>%
  select(microbe, Taxonomy)

temp <- gather(rarprev, microbe, prev, 2:6374, factor_key=TRUE)

# check that no samps are lost #obs temp = temp2 
temp2 <- merge(temp, taxkey, by="microbe")

rarprev <- temp2 %>%
    mutate(domain = gsub("(d__\\w+).*", "\\1", Taxonomy),
           kingdom = ifelse(grepl("k__", Taxonomy),gsub(".*(k__\\w+).*", "\\1", Taxonomy), NA),
           phylum = ifelse(grepl("p__", Taxonomy),gsub(".*(p__\\w+).*", "\\1", Taxonomy), NA),
           order = ifelse(grepl("o__", Taxonomy),gsub(".*(o__\\w+).*", "\\1", Taxonomy), NA),
           class = ifelse(grepl("c__", Taxonomy),gsub(".*(c__\\w+).*", "\\1", Taxonomy), NA),
           family = ifelse(grepl("f__", Taxonomy),gsub(".*(f__\\w+).*", "\\1", Taxonomy), NA),
           genus = ifelse(grepl("g__", Taxonomy),gsub(".*(g__\\w+).*", "\\1", Taxonomy), NA),
           species = ifelse(grepl("s__", Taxonomy),gsub(".*(s__.*)", "\\1", Taxonomy), NA)) %>%
  # for our purposes here just select genus
    dplyr::select(sample, genus, prev)
```

summarize genus prev
```{r}
rarprev <- rarprev %>%
  # mutate(genus = gsub(" ", "_", genus)) %>%
  # mutate(genus = gsub("[", "_", genus, fixed=TRUE)) %>%
  # mutate(genus = gsub("]", "_", genus, fixed=TRUE)) %>%
  filter(!is.na(genus)) %>%
  group_by(sample, genus) %>%
  summarize(prev = sum(prev)) %>%
  mutate(prev = ifelse(prev > 0, 1, 0)) %>%
  rename("microbe" = genus) 

# long to wide
rarprev <- spread(rarprev, microbe, prev)  
```


```{r get names of microbes with nonzero prevalence}
testprev <- colSums(rarprev[,-1])
microbes <- names(subset(testprev, testprev > 0))
```



# Object for loop
```{r}
surv.in <- clin %>%
  rename("sample" = "file_id.BAM") %>%
  inner_join(rarprev) %>%
  mutate(smoking = ifelse(pack_years_smoked >= 10, 1, 0),
         smoking = ifelse(is.na(smoking), 0, smoking),
         age = age_at_diagnosis/365.25,
         days = coalesce(days_to_death, days_to_last_follow_up),
         vitalstatus = ifelse(vital_status == "Dead", 1, 0),
         survstage = ifelse(grepl("IV", ajcc_clinical_stage), "Stage IV", ajcc_clinical_stage)) %>%
  drop_na(days, vitalstatus)
```
# Check survival of covariates

Looks like the model is able to work with all of these covariates on their own. I'm a little worried about the stage not being significant though. Let's also check all the covariates together.
```{r}
covars <- c("smoking", "age", "survstage")

univ.cov <- lapply(covars, function(x) coxph(as.formula(paste0("Surv(days, vitalstatus) ~ ", x)), data = surv.in))
names(univ.cov) <- covars

lapply(univ.cov, function(x) summary(x))
```

```{r}
test.allcov <- coxph(as.formula(paste0("Surv(days, vitalstatus) ~ ", paste(covars, collapse = " + "))),
                     data = surv.in)
summary(test.allcov)
```


# Run survival

```{r get formulas to try}
covars.string <- paste(covars, collapse = " + ")
forms <- paste0("Surv(days, vitalstatus) ~ ", microbes, " + ", covars.string)
names(forms) <- microbes
```


```{r The big survival calculation}
multivar.cph.ls <- lapply(microbes, function(x) extract_coxph_results(forms[[x]], x, surv.in))
multivar.cph <- bind_rows(multivar.cph.ls)
```

# Make nice tables
```{r}
nsamps = nrow(surv.in)
mics.in.dat <- surv.in %>%
  select(microbes)
tally.mics <- colSums(mics.in.dat) %>%
  as.data.frame() %>%
  rownames_to_column(var = "term") %>%
  rename("samples.with.microbe" = ".") %>%
  mutate(samples.without.microbe = nsamps - samples.with.microbe,
         pass.prevalence.split = ifelse(samples.without.microbe > 10 & samples.with.microbe > 10, "Yes", "No"))
```

```{r}
multivar.withns <- multivar.cph %>%
  mutate(padj.FDR = p.adjust(p.value, "fdr")) %>%
  left_join(tally.mics) %>%
  arrange(desc(pass.prevalence.split), p.value)
```

```{r}
write.csv(multivar.withns, "../data/survival_multivar-microbes_genus.csv", row.names = F)
```

# Aggregating Alphapappilomavirus

```{r}
test <- colnames(rarprev)
test[grep("Alphapap", test)]

to.agg <- paste0("Alphapapillomavirus.", c(5,6,9,7,11))

agg.prev <- rarprev %>%
  dplyr::select('sample', to.agg) %>%
  gather(-sample, key = "contributor", value = "prev") %>%
  group_by(sample) %>%
  summarise(HPV.prev = max(prev))

table(agg.prev$HPV.prev)
write.csv(agg.prev, "../data/HPV-prevalence-per-sample.csv", row.names = F)
```


### filter to microbes present in at least 10 samples and p.value < 0.05

# read in previously constructed survival table
```{r}
surv <- read.csv("../data/survival_multivar-microbes.csv")
```

```{r}
surv.new <- surv %>%
    filter(samples.with.microbe > 9,
           p.value < 0.05) %>%
    mutate(padj.FDR = p.adjust(p.value, "fdr"))
```

```{r}
write.csv(surv.new, "../data/survival_multivar-microbes_genus_filtered.csv")
```


